---
title: "Diphtheria Reproduction Rate - estimation from the Nigeria diphtheria outbreak"
author: "R Kamadjeu et al"
date: '`r format(Sys.Date())`'
output:
  html_document:
    toc: yes
    df_print: paged
    number_sections: no
bibliography: Diphtheria.bib
csl: the-pan-african-medical-journal.csl
link-citations: true
nocite:
    - '@item1'
    - '@item2'
suppress-bibliography: false
lang: en-US
zotero: true
---

```{r setup, include=FALSE}
remotes::install_github("paleolimbot/rbbt")
library(rbbt)
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Loading the dataset and packages and performing data management

#packman is provides a way to seamlessly manage packages
pacman::p_load(
  rio,          
  here,         
  lubridate,    
  aweek,        
  incidence2,   
  i2extras,     
  stringr,      
  forcats,      
  RColorBrewer, 
  tidyverse,     
  janitor,
  data.table,
  R0,
  table1,
  outbreaks,
  incidence
)

Kano_linelist_complete <- import("KANO_LINE_08_JAN.xlsx")

#The variables names in the dataset were  change because non appropriate (it had space).
Kano_linelist <- import("KANO_NEW.xlsx")
#We pick only the clinically compatible and the lab confirmed.
Kano_linelist2<-Kano_linelist %>% filter(fclass=="Clin compatible" | fclass=="Epid linked" | fclass=="Lab confirmed")
line_list<-Kano_linelist %>% filter(fclass=="Clin compatible" | fclass=="Epid linked" | fclass=="Lab confirmed") %>% select(donset)
line_list$CountCase<-1
#We create a data frame

#We create a new dataset with incidence by day. We eliminate data with missing values for donset.
LineList<-line_list%>% filter(!is.na(donset)) %>% group_by(donset) %>% summarise(
    N = sum(CountCase)
)


LineList$dayNum  <- as.Date(LineList$donset) -min(as.Date(LineList$donset))+1
LineList$date_of_onset<-as.Date(LineList$donset)
# days without cases are marked 0
LineList2 <- cbind(c(rep(0,max(LineList$dayNum))),c(rep(0,max(LineList$dayNum))))
LineList2[,1] <- c(1:max(LineList$dayNum))
LineList2[LineList$dayNum,2] <- LineList$N
LineList2 <- as.data.frame(LineList2)
names(LineList2) <- c("dayNum","N")


```




### Data management


```{r}
#### Description of the diphtheria outbreak in Kano State
Kano_linelist_complete$CountCase<-1
Kano_linelist_lga<-Kano_linelist_complete %>% filter(fclass=="Clin compatible" | fclass=="Epid linked" | fclass=="Lab confirmed") %>%
  group_by(lga) %>%
  summarise(cases = sum(CountCase))%>%
  filter(cases>=2)
#We write as a file

```

We model the epidemic overall. 


```{r}

#We create two dataset based on the two picks of the outbreak
outbreak.all<-line_list
#specifically setting the dates
date.onset<-as.Date(outbreak.all$donset)
#modelling the incidence of the entire outbreak over the 470 days (August 18, 2022, and 29 Nov 2023).
outbreak.Incidence.all.1 <- incidence::incidence(date.onset, interval = 1)
outbreak.Incidence.all.1

#finding a split
Outbreak.all.split<- incidence:::fit_optim_split(outbreak.Incidence.all.1)

Outbreak.all.split$split
Outbreak.all.split$fit

```

We find the date of peak incidence

```{r}

group.line.list<-line_list %>% mutate(CaseTag=1) %>% 
    group_by(donset) %>% 
  summarise(incidence=sum(CaseTag))

date.max.incidence<-group.line.list %>% filter(incidence == max(group.line.list$incidence))
#Date of max incidence and number of cases
date.max.incidence


```

From the date of pick incidence, we find the dates interval equal to approximately the outbreak doubling time (70 days), from date of pick incidence, until start of the outbreak

```{r}
# Load necessary package
library(lubridate)

# Define the start and end dates
end_date <- as.Date("2022-08-18")
start_date <- as.Date("2023-08-10")

# Initialize variables
current_date <- start_date
date_list <- c()

# Subtract 70 days repeatedly until the date is less than the start date
while (current_date > end_date) {
  current_date <- current_date - 70
  if (current_date >= end_date) {
    date_list <- c(date_list, as.character(current_date))
  }
}

# vector list of date limits used to split the phase 1 of the outbreak.
date_list


```


We get the time with highest r, for each of the 70 days interval

```{r}

#Vector dates identified above 
dates <- date_list

# Initialize a vector to store differences
r_interval <- numeric(length(dates) - 1)
r_cumulative <- numeric(length(dates) - 1)

# Loop through the vector to calculate differences
for (i in 1:(length(dates) - 1)) {
  date1 <- dates[i]
  date2 <- dates[i + 1]
  print(i)
  print(date2)
  print(date1)
  outbreak.interval<-line_list%>% filter(between(donset, as.Date(date2), as.Date(date1)))
  print(nrow(outbreak.interval)) 
  date.var <- as.Date(outbreak.interval$donset)
  outbreak.Incidence.int <- incidence::incidence(date.var, interval = 1)
  #r_cumulative[i]<-sum(i.1.outcome.c$counts)
  model.fit <- fit(outbreak.Incidence.int)
  #r.model<-model.fit$info$r
  r_interval[i] <- model.fit$info$r
}
 #r_cumulative[i]
# Print the differences, identify r for each time intervals
r_interval


```


```{r}
#fiiting the phase of mamximum growth
dat<-as.Date(line_list$donset)
class(dat)
incidence.all.1<-incidence(dat,interval=1)
best.fit <- fit_optim_split(incidence.all.1)
#plot(incidence.all.1)
max.fit <- fit(incidence.all.1[287:357]) #"2023-08-10", "2023-06-01"

```

#Modeling the interval of maximum exponential growth 

```{r}

  date1 <-  "2023-08-10"
  date2 <- "2023-06-11"
  outbreak.interval<-line_list%>% filter(between(donset, as.Date(date2), as.Date(date1)))
  date.var <- as.Date(outbreak.interval$donset)
  outbreak.Incidence.int <- incidence::incidence(date.var, interval = 1)
 # outbreak.Incidence.int.split<- incidence:::fit_optim_split(outbreak.Incidence.int)
#  r_interval <- outbreak.Incidence.int.split$fit$before$info$r
#  r_interval
  #plot(outbreak.Incidence.int)
  max.growth.early.fit <- fit( outbreak.Incidence.int)
  #plot(max.growth.early.fit)
  

```


Plotting the combine epi curve

```{r}

#Example, plotting the combined epi curve.

#value of r [1]  0.070505456 -0.008121465 -0.022075479  0.012369191  0.000000000

EpiCurve.all.split.2<-plot(incidence.all.1, fit = best.fit$fit, color = "red") %>%
  add_incidence_fit(Outbreak.all.split$fit) +
  #add_incidence_fit(max.fit)+
 geom_vline(xintercept = as.Date("2023-06-01"), color="blue", linewidth=0.5, linetype="dashed")+
 geom_vline(xintercept = as.Date("2023-03-23"), color="blue", linewidth=0.5, linetype="dashed")+
 geom_vline(xintercept = as.Date("2023-01-12"), color="blue", linewidth=0.5, linetype="dashed")+
 geom_vline(xintercept = as.Date("2023-08-10"), color="black", linewidth=0.5, linetype="dashed")+ #max incidence
 geom_vline(xintercept = as.Date("2022-11-03"), color="blue", linewidth=0.5, linetype="dashed") +
 geom_vline(xintercept = as.Date("2022-08-25"), color="blue", linewidth=0.5, linetype="dashed")+
  geom_vline(xintercept = as.Date("2023-08-27"), color="grey", linewidth=1, linetype="solid")+
 geom_line(data = max.fit$info$pred, aes(x=dates,y=fit),stat="identity",linetype = "dashed",linewidth=0.5)+
 geom_line(data = max.fit$info$pred, aes(x=dates,y=fit),stat="identity",linetype = "solid",linewidth=1.4, color="blue")+
  #geom_line(data = max.fit$info$pred, aes(x=dates,y=lwr),stat="identity",linetype = "dashed",linewidth=0.5)+
  #geom_line(data = max.fit$info$pred, aes(x=dates,y=upr),stat="identity",linetype = "dashed",linewidth=0.5)+
  theme_bw()+
  annotate(geom="text", x=as.Date("2023-07-05"), y=105, label="r = 0.078",
              color="red",fontface = 'italic')+
  annotate(geom="text", x=as.Date("2023-04-25"), y=105, label="r = -0.008",
              color="black",fontface = 'italic')+  
  annotate(geom="text", x=as.Date("2023-02-15"), y=105, label="r = -0.022",
              color="black",fontface = 'italic')+  
  annotate(geom="text", x=as.Date("2022-12-10"), y=105, label="r = 0.012",
              color="black",fontface = 'italic')+    
annotate(geom="text", x=as.Date("2022-09-20"), y=105, label="r = 0.000",
              color="black",fontface = 'italic')+ 
  annotate(geom="text", x=as.Date("2023-02-15"), y=122, label="Phase 1",
              color="darkgrey",fontface = 'italic')+ 
  annotate(geom="text", x=as.Date("2023-10-15"), y=122, label="Phase 2",
              color="darkgrey",fontface = 'italic')+   
annotate(
        'curve',
         x = as.Date("2022-08-18"), 
         y = 118, 
         xend = as.Date("2023-08-27"), 
         yend = 118, 
         linewidth=0.5,
         curvature = 0,
  arrow = arrow(
  type = "closed",
  ends = "both"
  ), 
  color="grey"
  ) +
annotate(
        'curve',
         x = as.Date("2023-08-27"), 
         y = 118, 
         xend = as.Date("2023-11-29"), 
         yend = 118, 
         linewidth=0.5,
         curvature = 0,
  arrow = arrow(
  type = "closed",
  ends = "both"
  ), 
  color="grey"
  ) +  
theme(
#    axis.text.x = element_text(color = "grey20", size = 15, ),
#    axis.text.y = element_text(color = "grey20", size = 15),
#    legend.text=element_text(size=15),
#    axis.title=element_text(size=15),
#    panel.background = element_blank(),
    panel.grid.major = element_line(colour = "gray80",linewidth=0.1),
    panel.grid.minor = element_line(colour = "gray80",linewidth=0.1),
#  
) +
  labs(
     y = "Daily incidence (N)",
     x = "Date of onset"
  )
EpiCurve.all.split.2


```


```{r, warning=FALSE}

library(outbreaks)
#We create two dataset based on the two picks of the outbreak
outbreak<-line_list
dat0 <- as.Date(outbreak$donset)
outbreak.Incidence.1 <- incidence::incidence(dat0, interval = 1)
outbreak.Incidence.1

Outbreak.split<- incidence:::fit_optim_split(outbreak.Incidence.1)

Outbreak.split$split
Outbreak.split$fit

```


calculating serial interval



```{r, warning=FALSE, echo=FALSE}
#We get the lgas
first.lap<-line_list

#cases.by.lga<-line_list %>% 
#  group_by(lga) %>% summarise(
#  caseCount  = n()
#)

cases.by.lga.lap1<-first.lap %>%group_by(lga) %>% summarise(
  caseCount  = n()
)

#head(arrange(cases.by.lga, desc(caseCount)), 10)

#getting the vector of LGAs with more than 2 cases. 
Lga.high.cases<-cases.by.lga.lap1%>%filter(caseCount>2)
#Create a vector for LGA with more than 2 cases
Selected.lga<-Lga.high.cases$lga
#Set the data frame that will hold the generation time
columns= c("Diff","lga") 
Gen.Time.dist.df = data.frame(matrix(nrow = 0, ncol = length(columns))) 
colnames(Gen.Time.dist.df) = columns
#dim(Gen.Time.dist.df)
#First, we select the specific LGA
#We create a function that generates, for each LGA, the generation time distribution
#We create a data.frame of dates of onset sorted by donset asc
for (value in Selected.lga) {
     gen.time.lga<-first.lap %>% filter(lga==!!enquo(value)) %>% arrange(donset) %>%
          group_by(donset) %>% summarise(
          date.count = n()
        )
    
    gen.time.lga.dis<-gen.time.lga %>%
        mutate(Diff = donset - lag(donset))%>% filter(!is.na(Diff))
    ##we had that distribution into the distribution
    diff.only<-gen.time.lga.dis$Diff #Create the vector
    Gen.Time.dist.df <- rbind(Gen.Time.dist.df, 
                      data.frame(
                        diff = diff.only,
                        lga = value
                        )) #bind the vector
}
 
 #Verify
#we create a numeric from days
Gen.Time.dist.df$lag<-as.numeric(Gen.Time.dist.df$diff)
#dim(Gen.Time.dist.df) 
Gen.Time.dist.df.2<-Gen.Time.dist.df %>% filter(lag<=21)
#summary(Gen.Time.dist.df.2$lag)
serial.int.dist <- ggplot(Gen.Time.dist.df.2, aes(x = lag, y = after_stat(density))) +
    geom_histogram(fill = "lightgrey", color = "lightgrey", binwidth = 2)+
    geom_density(color = "blue", linewidth = 2, bw=1, adjust=1)+
  geom_vline(xintercept = 2.8, colour="red", linetype="dashed", linewidth=0.75)+
    #geom_vline(xintercept = 3.48, colour="red", linetype="solid", linewidth=0.75)+
  theme_bw()+
   labs(                            # add plot labels
    x = "Time (days)",
    y = "Density",
    )+theme_bw()

serial.int.dist

#lag time is the serial interval distribution between dates of onset of symptoms of confirmed cases
lag.time<-Gen.Time.dist.df$lag
est.GT(serial.interval=lag.time)


```



*Reproduction numbers based on the interval of exponential growth from day

```{r, echo=FALSE, warning=FALSE, message=FALSE}

t1<-min(as.Date(line_list$donset))+288
t2<-min(as.Date(line_list$donset))+358 #max incidence
var_mean_si<-X
var_sd_si<-Y
GT.Diph <- generation.time("lognormal", c(var_mean_si, var_sd_si))  
#We create a vector of cases
#The data selected should be from the point the growth of the epidemic is exponential
DiphData<-line_list %>% filter(donset>=t1 & donset<=t2) %>%group_by(donset) %>% summarise(
  N = n()
)

DiphData2<-line_list%>%group_by(donset) %>% summarise(
  N = n()
)
#Estimating the generation time from that interval. 
    gen.time.interval<-line_list %>%
        dplyr::mutate(Diff = donset - lag(donset)) %>% dplyr::filter(!is.na(Diff))
    ##we had that distribution into the distribution
    diff.only.interval<-gen.time.interval$Diff #Create the vector
    
Dates<-as.Date(DiphData$donset)
Number<-DiphData$N
ProcessData <- as.data.frame(cbind(Dates, Number))
names(ProcessData) <- c("dayNum","N")
ProcessData$dayNum <- ProcessData$dayNum-min(ProcessData$dayNum)+1

res.EG.1 <- estimate.R(Number, GT=GT.Diph, methods=c("EG"))
res.ML.1 <- estimate.R(Number, GT=GT.Diph, methods=c("ML"))
res.SB.1 <- estimate.R(Number, GT=GT.Diph, methods=c("SB"))
res.TD.1 <- estimate.R(Number, GT=GT.Diph, methods=c("TD"))
#display results
res.EG.1
res.ML.1
res.SB.1
res.TD.1
#display min, mean and median for methods
mean(res.SB.1$estimates$SB$R)
min(res.SB.1$estimates$SB$R)
median(res.SB.1$estimates$SB$R)
mean(res.TD.1$estimates$TD$R)
min(res.TD.1$estimates$TD$R)
median(res.TD.1$estimates$TD$R)




```

Estimating sample mean and 95% CI of the Rt distribution obtained with method SB


```{r}
library( boot)
outPut.SB<-estimate.R(Number, GT=GT.Diph, methods=c("SB"),begin=1,end=69)
#We extract GT obtained from SB method into a vector
Data.SB<-outPut.SB$estimates$SB[1]$R
hist(Data.SB) #Quite a non.normally distributed sample
#Obtaining the boot mean
Bmean <- function(data, indices) {
  d <- data[indices] # allows boot to select sample 
    return(mean(d))
}

# bootstrapping with 1000 replications 
results.SB <- boot(data=Data.SB, statistic=Bmean, R=1000)
results.SB 
plot(results.SB)


# get 95% confidence interval 
boot.ci(results.SB, type=c("norm", "basic", "perc", "bca"))


```


Bootstrap TD

```{r}
outPut.TD<-estimate.R(Number, GT=GT.Diph, methods=c("TD"),begin=1,end=67)
Data.TD<-outPut.TD$estimates$TD[1]$R
hist(Data.TD) 
# bootstrapping with 1000 replications 
results.TD <- boot(data=Data.TD, statistic=Bmean, R=1000)
results.TD 
plot(results.TD)

boot.ci(results.TD, type=c("norm", "basic", "perc", "bca"))

```


```{r}

sensitivity.analysis(ProcessData$N, GT.type="lognormal", GT.mean=seq(1,6,1), GT.sd.range=1, begin=1, end=68, est.method="ML", sa.type="GT")

```


```{r}

res.R <- estimate.R(ProcessData$N, GT=GT.Diph, methods=c("EG","ML","SB","TD"))
res.R
#jpeg("rplot.jpg", width = 350, height = 350)
#plot.sa<-plot(res.R)
#dev.off()
#plotfit(res.R)
```
Sensitivity analysis


```{r}

sensitivity.analysis(ProcessData$N, GT.type="lognormal", GT.mean=seq(1,5,1), GT.sd.range=1, begin=1, end=68, est.method="EG", sa.type="GT")

```

```{r}

Inspect<-inspect.data(ProcessData$N, GT =GT.Diph)
Inspect

```


```{r}

sa.time( ProcessData$N, GT=GT.Diph, begin = 1, end = 68, est.method="EG", t = NULL, date.first.obs = NULL)
```

```{r}

tmp.ml<-sensitivity.analysis(sa.type = "GT", incid = ProcessData$N, GT.type = "lognormal", GT.mean = seq(1,6,1), GT.sd.range = 1, begin = 1, end = 67, est.method = "ML")
tmp.eg<-sensitivity.analysis(sa.type = "GT", incid = ProcessData$N, GT.type = "lognormal", GT.mean = seq(1,6,1), GT.sd.range = 1, begin = 1, end = 67, est.method = "EG")
tmp.ml
tmp.eg

tmp.ml.df<-as.data.frame(tmp.ml)
tmp.eg.df<-as.data.frame(tmp.eg)

tmp.ml.df<-tmp.ml.df %>% mutate(
  method = "ML"
)
tmp.eg.df<-tmp.eg.df %>% mutate(
  method = "EG"
)

tmp.bind.sens<-rbind(tmp.ml.df,tmp.eg.df)

tmp.bind.sens<-tmp.bind.sens %>% mutate(
  GT.Mean2 = round(as.double(GT.Mean), digits = 2),
  R2 = round(as.double(R), digits = 2),
  CI.lower2 = round(as.double(CI.lower), digits = 2),
  CI.upper2 =round(as.double(CI.upper), digits = 2),
)

```

Plotting the sensitivity analysis


```{r}


pd1 = position_dodge(0.4)

ggplot(tmp.bind.sens, aes(x = GT.Mean2, y = R2, fill=method, color=method))+
  geom_point(alpha=0.7) +
  #stat_summary(fun.data=mean_cl_boot, geom="errorbar", width=5, position=pd1) +
  #stat_summary(fun.y=mean, geom="point", size=2, position=pd1) +
  #scale_fill_manual(values=hcl(c(15,195), 100, 60)) +
 # scale_color_manual(values=hcl(c(15,195), 50, 40)) +
  theme_bw()


pd1 = position_dodge(0.2)
fp <- ggplot(data=tmp.bind.sens, aes(x=GT.Mean2, y=R2, ymin=CI.lower2, ymax=CI.upper2,fill=method, color=method)) +
     geom_line( color="grey") +
        geom_pointrange(size=1.25,position=pd1,shape = 21, stroke=2) + 
       # coord_flip() +  # flip coordinates (puts labels on y axis)
        xlab("Mean Generation time (days)") + ylab("R0 (95% CI)") +
        theme_bw()+
    #theme(legend.position = 'top') 
    theme(
      legend.position = c(0.2,0.8),
      axis.text.x = element_text(color = "grey20", size = 15, ),
      axis.text.y = element_text(color = "grey20", size = 15),
      legend.text=element_text(size=15),
      axis.title=element_text(size=15),
    )
fp 
#ggsave("fp.jpeg",fp, dpi = 300)

```

Testing the vector for plotting individual methods

```{r}
time<-seq(1,67,1)
observed<-Number[-length(Number)]
predicted.EG.1<-res.EG.1$estimates$EG$pred
predicted.ML.1<-res.ML.1$estimates$ML$pred
predicted.SB.1<-res.SB.1$estimates$SB$pred
predicted.TD.1<-res.TD.1$estimates$TD$pred


df.EG1=data.frame(time, observed, predicted.EG.1,predicted.ML.1,predicted.SB.1,predicted.TD.1) 


```

